---
title: "Starting with data"
author: "Data Carpentry contributors"
minutes: 20
---

```{r, include=FALSE, purl=FALSE}
source("setup.R")
```

------------

> ### Learning Objectives
>
> * Load external data from a .csv file into a data frame.
> * Describe what a data frame is.
> * Summarize the contents of a data frame.
> * Extract a vector from a data frame
> * Describe what a factor is.
> * Convert between strings and factors.
> * Reorder and rename factors.
> * Format dates.

------------

## Presentation of the Survey Data

```{r, echo=FALSE, purl=TRUE}
### Presentation of the survey data
```

We are studying the species repartition and weight of animals caught in plots in our study
area. The dataset is stored as a comma separated value (CSV) file.
Each row holds information for a single animal, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxon            | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from Figshare, and we will use `read_csv()` to
load into memory the content of the CSV file as an object of class `data.frame`. 
Inside the download.file command, the first entry is a character string with the source URL 
("https://ndownloader.figshare.com/files/2292169"). This source URL downloads a CSV file from 
figshare. The text after the comma ("data_raw/portal_data_joined.csv") is the destination of the 
file on your local machine. You'll need to have a folder on your machine called "data_raw" where 
you'll download the file. So this command downloads a file from Figshare, names it 
"portal_data_joined.csv," and adds it to a preexisting folder named "data_raw".

```{r, eval=FALSE, purl=TRUE}
download.file(url="https://ndownloader.figshare.com/files/2292169",
              destfile = "data_raw/portal_data_joined.csv")
```

If we haven't already done so, we can type `install.packages("tidyverse")` straight into the console. In fact, it's better to write this in the console than in our script for any package, as there's no need to re-install packages every time we run the script.

Then, to load the package type:

```{r, message = FALSE, warning = FALSE, purl = FALSE}
## load the tidyverse packages
library(tidyverse)
```

You are now ready to load the data:

```{r, results = 'hide', message = FALSE, purl = FALSE}
surveys <- read_csv("data_raw/portal_data_joined.csv")
```

This statement doesn't produce any output because, as you might recall,
assignments don't display anything. If we want to check that our data has been
loaded, we can see the contents of the data frame by typing its name: `surveys`.

In RStudio's `Environment` pane, this `Data` object can now be inspected by
clicking its name and the round `>` button left of it. The latter provides an
overview of all columns/variables grouped by data type/class: `character` and
`numeric`.

> ### Note
>
> `read_csv` assumes that fields are delineated by commas, however, in several
> countries, the comma is used as a decimal separator and the semicolon (`;`) is
> used as a field delineator. If you want to read in this type of files in R,
> you can use the `read_csv2` function. It behaves exactly like `read_csv` but
> uses different parameters for the decimal and the field separators. If you are
> working with another format, they can be both specified by the user. Check out
> the help for `read_csv()` by typing `?read_csv` to learn more. There is also the `read_delim()` for 
> in tab separated data files.
>
> In addition to the above versions of the csv format, you should develop the habits
> of looking at and record some parameters of your csv files. For instance, 
> the character encoding, control characters used for line ending, date format
> (if the date is not splitted into three variables), and the presence of unexepected
> [newlines](https://en.wikipedia.org/wiki/Newline) are important characteristics of your data files.
> Those parameters will ease up the import step of your data in R.

Using the columns, we're going to learn some common analysis functions later:

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statistics on grouped data
- `arrange()`: sort results
- `count()`: count discrete values

But first, some theory.

## What are data frames (and tibbles)?

Data frames are the _de facto_ data structure for most tabular data, and what we
use for statistics and plotting. Tibbles are the tidyverse's modern variant.
Essentially, they are a bit nicer to work with.

Both can be created by hand, but most commonly they are generated by the
functions `read...` functions; in other words, when importing
spreadsheets from your hard drive (or the web).

A data frame / tibble is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because columns are
vectors, each column must contain a single type of data (e.g., characters, integers,
factors). For example, here is a figure depicting a data frame comprising a
numeric, a character, and a logical vector.

![](./img/data-frame.svg)


We can see this when inspecting the <b>str</b>ucture of a data frame
with the function `str()`:

```{r, purl=FALSE}
str(surveys)
```

## Inspecting `data.frame` Objects

We already saw how calling `surveys` itself or `str(surveys)` provided us with 
content and the structure of the data frame / tibble. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data. Let's try them out!

* Size:
    * `dim(surveys)` - returns a vector with the number of rows in the first element,
          and the number of columns as the second element (the **dim**ensions of
          the object)
    * `nrow(surveys)` - returns the number of rows
    * `ncol(surveys)` - returns the number of columns

* Content:
    * `head(surveys)` - shows the first 6 rows
    * `tail(surveys)` - shows the last 6 rows

* Names:
    * `names(surveys)` - returns the column names (synonym of `colnames()` for `data.frame`
	   objects)
    * `rownames(surveys)` - returns the row names

* Summary:
    * `str(surveys)` - structure of the object and information about the class, length and
	   content of  each column
    * `summary(surveys)` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.


> ### Challenge
>
> Based on the output of `str(surveys)`, can you answer the following questions?
>
> * What is the class of the object `surveys`?
> * How many rows and how many columns are in this object?
> * How many species have been recorded during these surveys?
>
> ```{r, answer=TRUE, results="markup", purl=FALSE}
>
> str(surveys)
>
> ## * class: data frame
> ## * how many rows: 34786,  how many columns: 13
> ## * how many species: 48
>
> ```


```{r, echo=FALSE, purl=TRUE}

## Challenge
## Based on the output of `str(surveys)`, can you answer the following questions?
## * What is the class of the object `surveys`?
## * How many rows and how many columns are in this object?
## * How many species have been recorded during these surveys?

```



## Extracting vectors from data frames

```{r, echo=FALSE, purl=TRUE}

## Extracting vectors from data frames
```

Our survey data frame has rows and columns (its 2 dimensions). If we want to
extract some specific column we can do:

```{r, eval = FALSE, purl=FALSE}
surveys$species_id
```

RStudio conveniently supports this `$`-notation by autocompletion, which is the
reason why column names should not contain spaces. If you have a dataset that
does, use `dataset[["column name"]]`.
